---
title: "NDVI_poly_report"
output: 
  html_document: 
    fig_caption: true
date: "2025-05-01"
editor_options: 
  chunk_output_type: console
---
## PRI model report
```{r setup, include=FALSE}
library(here)
setwd(here())
source('./Scripts/.PACKAGES.R')
mod <- readRDS('./NDVI.rds')
index <- read.csv('./Data/Index.csv') %>%
  filter(branch == 'top') %>% filter(days < 59)

data <- index
data$treatment <- as.factor(data$treatment)
data$clone <- factor(data$clone, labels = c('c1', 'c2','c3','c4','c5','c6'))
contrasts(data$clone) <- contr.sum(levels(data$clone))
data$days_scaled <- drop(scale(data$days))

posterior_samples <- posterior_samples(mod)

```

Posterior predictive checks allow you to compare the observed data to data simulated from the model. This can help you assess how well the model captures the data.

```{r}
pp_check(mod)
```
Trace plots help you visualize the convergence of the Markov Chain Monte Carlo (MCMC) chains. You can use the `plot()` function to generate trace plots for the parameters.

```{r}
# traceplots 
plot(mod)
```
Rhat values (also known as the potential scale reduction factor) indicate whether the chains have converged. Values close to 1 suggest good convergence. You can check Rhat values in the model summary.
```{r}
# Check Rhat values
rhat_values <- summary(mod)$fixed[, "Rhat"]
print(rhat_values)
# Random effects diagnostics
#ranef(mod)
```
Divergent transitions can indicate problems with the model specification or the priors. You can check for divergent transitions using the summary() function or by plotting the model.
```{r}
# Check for divergent transitions
summary(mod)$diagnostics
```

the fitted v modeled predictions
```{r}
# Visualize predictions
predictions <- fitted(mod)
plot(data$PRI, predictions[,1], main = "Observed vs Fitted", xlab = "Observed", ylab = "Fitted")
abline(0, 1, col = "red")  # Add a 45-degree line
```

```{r, fixed effects}
fixed_effects <- posterior_samples %>%
  select(!starts_with("b_")) %>%  # Select all fixed effects (starts with "b_")
  pivot_longer(cols = everything(), names_to = "effect", values_to = "estimate") 

sum_mod <- summary(mod)
sum_df <- as.data.frame(as.data.frame(sum_mod$fixed)) 
sum_df$noEffect <- (sum_df$`l-95% CI` <= 0 & sum_df$`u-95% CI`>=0)
print(sum_df)  
```



```{r include=FALSE}
em_week <- emmeans(
  mod, 
  specs = pairwise ~ treatment | days_scaled,
  at = list(days_scaled = seq(min(data$days_scaled), 
                              max(data$days_scaled), 
                              length = 8))  # Evaluates at original weeks
)

em_clone_week <- emmeans(
  mod, 
  specs = pairwise ~ treatment | days_scaled | clone,
  at = list(days_scaled = seq(min(data$days_scaled), 
                              max(data$days_scaled), 
                              length = 8))  # Evaluates at original weeks
)
```









# plots over time 
```{r}
# Plot treatment differences
plot(em_week$contrasts) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Drought vs. Control: Weekly Contrasts",
       y = "PRI Difference (Drought - Control)")
```
Figure: drought effect over time

```{r}
contrasts_df <- as.data.frame(em_clone_week$contrasts)
ggplot(contrasts_df, aes(x = days_scaled, y = estimate, color = contrast)) +
  geom_point() +
  geom_line() +
  geom_errorbar(aes(ymin = lower.HPD, ymax = upper.HPD), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Drought vs. Control: Weekly Contrasts by Clone",
       y = "PRI Difference (Drought - Control)",
       x = "Days Scaled") +
  facet_wrap(~ clone) +  # Facet by clone
  theme_minimal() +
  theme(legend.position = "bottom")
```
Figure: drought effect over time by clone

